Future challenges for Research Software Engineering

### 2025-09-26 correspondence/rebuttal-1.md

Now it is lower-case, except in names or publication titles.

> The titles in section 2 aren’t consistent in how they are capitalised

Thanks for spotting this, now the titles are properly capitalised.

> Page 2
> In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these “Goal” boxes. – The use of the term “Achieved Goals” is a bit confusing.  I understand it’s meant that these would have been achieved by 2035, but then the text under these boxes talks about 2035 as a future state. I’d suggest rewording these boxes to “2035 Goals’ or similar.
> Page 3
> Story 1: In order to give a more concrete setting, imagine a world ten years from now. Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred. – These are both incomplete sentences, ie it's a group of words that doesn't express a complete thought and is missing a necessary element, like a subject or a verb. 
> Page 4
> RSEng-relevant open educational resources and Master’s programmes are tracked in the Learning and Teaching RSE database2 and UK SSI resources hub3. The Carpentries now offer specialised software workshops for data scientists4, librarians5, and soon HPC practitioners6. – This seems to be about the future state, but it’s not clear what “now” means here (probably 2035?) – for example, a few paragraphs earlier there is wording “This means, that today’s workshop content of proper versioning and data management could become obsolete by using more intuitive tools.” – where today means 2025. Can it be made clearer throughout the paper when the text is talking about 2025 versus 2035?

### 2025-09-26 correspondence/rebuttal-1.md

> 1. The authors do not mention Germany as the focus of their discussion, probably because they want to address a broader aspect of RSEng that could apply to other countries or even globally. However, the history, focus and references clearly indicate a German perspective. I suggest either making it clear that the vision and discussion are limited to Germany, or discussing why the focus of the paper is rather nationalistic yet applies to a broader range of locations, and explaining where and why the stated aspects differ.

Thank you for highlighting this. We have added some text to the second paragraph of the introduction highlighting that with the paper focusing on the outputs of the German RSE conference, the thoughts and perspectives presented are largely German-focused. However, we also note that this work has broader relevance and we aim to highlight resources from the wider community where relevant.

> 2. The authors switch between the past and present tenses, which makes reading the article difficult. They sometimes address present tools, infrastructure and activities, and sometimes write as if future tools, infrastructure and activities have been made available or developed. For example, the "Achieved Goals" refer to goals achieved by 2035, but the tenses change throughout the boxes. "RSEng-relevant open educational resources..." Some resources are available and some are not; however, the style of writing makes it difficult to tell the difference. I would propose that the authors use tenses to clearly distinguish between past, present and future activities, tools and resources, and not write as if future activities have already been successful. Additionally, the authors write about their hopes, which seem off-topic for a report or opinion paper. I would suggest not writing about hopes at all. Future trend statements should be based on knowledge.

Thanks for spotting. We have used stronger language in the revision. We have also passed over the document to make the tenses homogeneous.

@@

### Suggestions

> It might be useful to note that this paper on RSE in 2030 exists https://arxiv.org/pdf/2308.07796, although the workshop participants didn’t know about it so it wasn’t incorporated in their thinking.

> Reviewer C also uploaded a file with more feedback

> Page 2
> In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these “Goal” boxes. – The use of the term “Achieved Goals” is a bit confusing.  I understand it’s meant that these would have been achieved by 2035, but then the text under these boxes talks about 2035 as a future state. I’d suggest rewording these boxes to “2035 Goals’ or similar.

We have reworded them to RSEng in 2035.

> Page 3
> Story 1: In order to give a more concrete setting, imagine a world ten years from now. Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred. – These are both incomplete sentences, ie it's a group of words that doesn't express a complete thought and is missing a necessary element, like a subject or a verb.

We have reworded this sentence in order to provide more natural british english.

> Page 4
> RSEng-relevant open educational resources and Master’s programmes are tracked in the Learning and Teaching RSE database2 and UK SSI resources hub3. The Carpentries now offer specialised software workshops for data scientists4, librarians5, and soon HPC practitioners6. – This seems to be about the future state, but it’s not clear what “now” means here (probably 2035?) – for example, a few paragraphs earlier there is wording “This means, that today’s workshop content of proper versioning and data management could become obsolete by using more intuitive tools.” – where today means 2025. Can it be made clearer throughout the paper when the text is talking about 2025 versus 2035?

Since the inconsistent use tof times was mentioned by multiple reviewers, we went over the paper multiple times to be more consistent with regard to the use of tenses.

### 2025-09-26 future_challenges_for_RSEng.tex
fix typo
provide answers to some reviewer feedback

\begin{story}{Introduction}{intro}
In order to give a more concrete setting, imagine a world ten years from now 
which is not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
were able to ride the waves of their career in RSEng and are now pondering early retirement on a sun blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems that have been solved by then?
Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
And what are the issues that have newly emerged during this time?
In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
of their future in respective small story boxes.
\end{story}

@@

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This is supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of specialised RSEs with RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally, researchers interact seamlessly with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

>>>>

\begin{story}{Introduction}{intro}
To provide a more concrete setting, imagine a world ten years from now --
not too overwhelmingly far into the future, but distant enough for some changes to have occurred.
Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
were able to ride the waves of their career in RSEng and are now pondering early retirement on a sun blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems that have been solved by then?
Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
And what are the issues that have newly emerged during this time?
In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
of their future in respective small story boxes.
\end{story}

@@

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This is supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of specialised RSEs with RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally7, researchers interact seamlessly with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
